---
title: "README"
author: "Zachary McCaw"
date: "`r Sys.Date()`"
output: 
  html_document: 
    keep_md: TRUE
--- 

# Functions for Survival Analysis

Zachary R. McCaw <br>
Updated: 2023-09-26

```{r global_options, include=FALSE}
knitr::opts_chunk$set(cache = FALSE, fig.align = "center")
```

```{r}
suppressPackageStartupMessages({
  library(dplyr)
  library(SurvUtils)
})
```

# Installation

```{r, eval=FALSE}
devtools::install_github(repo = "zrmacc/SurvUtils")
```

# Data Generation

Generates survival data with exponential event times and censoring. Optionally, the subject-specific event rate may depend on a set of covariates and/or a gamma-frailty.

```{r}
data <- SurvUtils::GenData(
  base_event_rate = 1.0,
  censoring_rate = 0.25,
  n = 100,
  tau = 4.0
)
head(data)
```

# Estimation

## One Sample

### Kaplan-Meier

* Tabulates the cumulative hazard and survival functions, along with variance estimates and confidence intervals.

```{r}
km_tab <- SurvUtils::TabulateKM(data)
head(km_tab)
```

#### Influence function

The influence function expansion of the Kaplan-Meier estimator is:

$$
\sqrt{n}\big\{ \hat{S}(t) - S(t) \big\} = \frac{1}{\sqrt{n}}\sum_{i=1}^{n}\psi_{i}(t) + o_{p}(1)
$$
where:

$$
\psi_{i}(t) = -S(t) \int_{0}^{t}\frac{dM_{i}(u)}{n^{-1} Y(u)}
$$
and $dM_{i}$ is the counting-process martingale:

$$
dM_{i}(u) = dN_{i}(u) - Y_{i}(u)dH(u)
$$

```{r}
n <- 1000
data <- GenData(n = n)
tau <- 1.0

# Influence function calculation.
influence <- KMInfluence(data, tau = tau)

# Estimated variance of the KM estimator at time tau.
inf_var <- mean(influence^2) / n
inf_se <- sqrt(inf_var)
print(round(inf_se, digits = 4))
```

### Event Rate, Percentile, Restricted Mean Survival

* Calculate the event rate at a point in time.

```{r}
# Rate.
SurvUtils::OneSampleRates(data, tau = 1.0)
```

```{r}
# Percentile: median.
SurvUtils::OneSamplePercentiles(data, p = 0.5)
```

```{r}
# RMST.
SurvUtils::OneSampleRMST(data, tau = 1.0)
```

## Two Sample

### Generate Data
```{r}
data0 <- SurvUtils::GenData(
  base_event_rate = 1.0,
  censoring_rate = 0.25,
  n = 100,
  tau = 4.0
)
data0$arm <- 0

data1 <- SurvUtils::GenData(
  base_event_rate = 0.5,
  censoring_rate = 0.25,
  n = 100,
  tau = 4.0
)
data1$arm <- 1
data <- rbind(data0, data1)
```

### Compare Rates
```{r}
SurvUtils::CompareRates(data, tau = 1.0)
```

### Compare RMSTs
```{r}
SurvUtils::CompareRMSTs(data, tau = 1.0)
```

### Compare Cox Models
Compare the predictive performance of Cox models based on different sets of covariates with respect to their c-statistics on held-out data via k-fold cross validation.
```{r}
# Simulate data.
n <- 1000
x1 <- rnorm(n)
x2 <- rnorm(n)
data <- SurvUtils::GenData(
  covariates = cbind(x1, x2),
  beta_event = c(1.0, -1.0)
)

# Evaluate.
eval <- CompreCoxCstat(
  status = data$status,
  time = data$time,
  x1 = data %>% dplyr::select(x1, x2),
  x2 = data %>% dplyr::select(x1)
)

head(round(eval, digits = 3))
```

# Plotting

```{r, cache=TRUE}
# Generate data.
arm1 <- SurvUtils::GenData(base_event_rate = 0.8)
arm1$arm <- 1
arm0 <- SurvUtils::GenData(base_event_rate = 1.0)
arm0$arm <- 0
data <- rbind(arm1, arm0)
```

## One Sample

### Standard Kaplan-Meier

```{r, cache=TRUE, }
x_breaks <- seq(from = 0.0, to = 4.0, by = 0.50)
data0 <- data %>% dplyr::filter(arm == 0)
fit0 <- Temporal::FitParaSurv(data0)  # Optional parametric fit. 
q_km <- SurvUtils::PlotOneSampleKM(data0, fit = fit0, x_breaks = x_breaks, x_max = 4)
q_nar <- SurvUtils::PlotOneSampleNARs(data0, x_breaks = x_breaks, x_max = 4)
cowplot::plot_grid(
  plotlist = list(q_km, q_nar),
  align = "v",
  axis = "l",
  ncol = 1,
  rel_heights = c(3, 1)
)
```

### AUC

```{r, cache=TRUE, }
x_breaks <- seq(from = 0.0, to = 4.0, by = 0.50)
data0 <- data %>% dplyr::filter(arm == 0)
q_auc <- SurvUtils::PlotOneSampleAUC(data0, x_breaks = x_breaks, x_max = 4, tau = 3)
q_nar <- SurvUtils::PlotOneSampleNARs(data0, x_breaks = x_breaks, x_max = 4)
cowplot::plot_grid(
  plotlist = list(q_auc, q_nar),
  align = "v",
  axis = "l",
  ncol = 1,
  rel_heights = c(3, 1)
)
```

## Two Sample

```{r, cache=TRUE}
x_breaks <- seq(from = 0.0, to = 4.0, by = 0.50)
contrast <- Temporal::CompParaSurv(data)  # Optional parametric fit. 
q_km <- SurvUtils::PlotTwoSampleKM(data, contrast = contrast, x_breaks = x_breaks, x_max = 4)
q_nar <- SurvUtils::PlotTwoSampleNARs(data, x_breaks = x_breaks, x_max = 4)
cowplot::plot_grid(
  plotlist = list(q_km, q_nar),
  align = "v",
  axis = "l",
  ncol = 1,
  rel_heights = c(3, 1)
)
```